---
output-file: data-wrangling-and-visualization-of-big-data
execute:
  enabled: true
  echo: true
  warning: false
  message: false
fig-width: 9
fig-height: 5.5
fig-align: center
---

# Data Wrangling and Visualization of Big Data

## Overview

This chapter works with the [Google Community Mobility Reports](https://www.google.com/covid19/mobility/) dataset. The reports describe changes in visits to places such as retail and recreation venues, grocery stores, parks, transit stations, and workplaces during the COVID-19 pandemic.

The original worldwide CSV is more than 1 GB. To keep this lesson fast and reproducible, the repository includes a compact snapshot containing daily **state-level US observations** from February 2020 through October 2022. County records and the national aggregate are excluded. The script `scripts/prepare_google_mobility.py` documents how the snapshot was created from Google's archived US files.

::: {.callout-note title="How to interpret the values"}
A value of `-20` means activity was 20% below Google's baseline; `15` means activity was 15% above it. These values describe changes in movement. They are not case counts and should not be interpreted as causal effects.
:::

## Load packages

The tidyverse supplies tools for importing, transforming, and plotting data. We use lubridate for dates, cowplot for combining figures, and knitr for compact tables.

```{r}
#| label: setup

library(tidyverse)
library(lubridate)
library(cowplot)
library(knitr)

theme_set(theme_minimal(base_size = 12))
```

## Read and inspect the data

Start by importing the local snapshot. `glimpse()` is especially useful for unfamiliar data because it shows dimensions, column types, and representative values without flooding the page.

```{r}
#| label: read-mobility-data

google_mobility <- read_csv(
  "data/google-mobility-us-states.csv",
  show_col_types = FALSE
)

glimpse(google_mobility)
```

Before analyzing anything, confirm the number of observations, geographic coverage, and date range. These checks catch incomplete downloads and unexpected filters early.

```{r}
#| label: tbl-data-summary
#| tbl-cap: "Google Mobility snapshot at a glance."

tibble(
  measure = c("Rows", "States and DC", "First observation", "Last observation"),
  value = c(
    format(nrow(google_mobility), big.mark = ","),
    n_distinct(google_mobility$sub_region_1),
    format(min(google_mobility$date)),
    format(max(google_mobility$date))
  )
) |>
  kable(col.names = c("Measure", "Value"))
```

## Add useful date and region features

R includes state names and Census region labels. We can turn those built-in vectors into a lookup table and join them to the mobility records. We also derive several date features that will support grouping at different time scales.

```{r}
#| label: prepare-mobility-data

state_lookup <- tibble(
  sub_region_1 = state.name,
  us_division = state.division,
  us_region = state.region,
  state_land_area = state.area
)

google_mobility_US <- google_mobility |>
  mutate(
    dt_month = floor_date(date, "month"),
    dt_year = year(date),
    dt_month_label = month(date, label = TRUE, abbr = TRUE),
    dt_weekday = wday(date, label = TRUE, abbr = TRUE),
    dt_weekend = dt_weekday %in% c("Sat", "Sun")
  )

google_mobility_US_w_lu <- google_mobility_US |>
  inner_join(state_lookup, by = "sub_region_1")
```

An explicit join key makes the relationship between the tables clear and prevents accidental joins on similarly named variables. The built-in state table does not include the District of Columbia, so the joined analysis contains the 50 states.

## Count records by state

Record counts help reveal uneven coverage. Here we show only the first and last five states alphabetically; printing all 50 would add length without adding much insight.

```{r}
#| label: tbl-state-records
#| tbl-cap: "Example record counts by state."

records_per_state <- google_mobility_US |>
  count(sub_region_1, name = "n_records") |>
  arrange(sub_region_1)

bind_rows(
  slice_head(records_per_state, n = 5),
  slice_tail(records_per_state, n = 5)
) |>
  kable(
    col.names = c("State", "Daily observations"),
    format.args = list(big.mark = ",")
  )
```

## Compare national trends

Daily state records are too dense for a first overview. Aggregating to monthly averages reduces noise and makes the broad temporal pattern easier to see. Reshaping the four place categories from wide to long format lets one ggplot specification draw all four series.

```{r}
#| label: create-national-summary

national_monthly <- google_mobility_US |>
  select(
    dt_month,
    retail_and_recreation_percent_change_from_baseline,
    parks_percent_change_from_baseline,
    workplaces_percent_change_from_baseline,
    residential_percent_change_from_baseline
  ) |>
  pivot_longer(
    -dt_month,
    names_to = "place_type",
    values_to = "percent_change"
  ) |>
  mutate(
    place_type = recode(
      place_type,
      retail_and_recreation_percent_change_from_baseline = "Retail & recreation",
      parks_percent_change_from_baseline = "Parks",
      workplaces_percent_change_from_baseline = "Workplaces",
      residential_percent_change_from_baseline = "Residential"
    )
  ) |>
  group_by(dt_month, place_type) |>
  summarise(
    percent_change = mean(percent_change, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
#| label: fig-national-trends
#| fig-cap: "Average monthly mobility change across US states. The dashed line marks the pre-pandemic baseline."

ggplot(national_monthly, aes(dt_month, percent_change, color = place_type)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_line(linewidth = 0.9) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    x = NULL,
    y = "Change from baseline (%)",
    color = NULL,
    title = "Mobility patterns changed sharply in early 2020"
  ) +
  theme(legend.position = "bottom")
```

The sharp drops in retail, recreation, and workplace activity are visible immediately. Residential activity moves in the opposite direction because people spent more time near home. Parks are more variable and seasonal, which is a reason to avoid summarizing the entire period with one average.

## Compare Census regions

National averages can hide geographic variation. Grouping states into Census regions strikes a useful balance: it preserves broad spatial differences without drawing 50 overlapping lines.

```{r}
#| label: create-regional-summary

regional_monthly <- google_mobility_US_w_lu |>
  group_by(us_region, dt_month) |>
  summarise(
    `Retail & recreation` = mean(
      retail_and_recreation_percent_change_from_baseline,
      na.rm = TRUE
    ),
    Parks = mean(parks_percent_change_from_baseline, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(
    c(`Retail & recreation`, Parks),
    names_to = "place_type",
    values_to = "percent_change"
  )
```

```{r}
#| label: fig-regional-trends
#| fig-cap: "Monthly retail and parks mobility by US Census region."

ggplot(regional_monthly, aes(dt_month, percent_change, color = us_region)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey65") +
  geom_line(linewidth = 0.85) +
  facet_wrap(~place_type, ncol = 1, scales = "free_y") +
  scale_color_brewer(palette = "Set1") +
  labs(
    x = NULL,
    y = "Change from baseline (%)",
    color = "Region"
  ) +
  theme(legend.position = "bottom")
```

Faceting gives each outcome its own vertical scale while retaining a shared time axis. That makes within-outcome regional comparisons easier, but it also means readers should not compare vertical distances across panels without checking the axes.

## Focus on selected states

Filtering is useful when the research question concerns specific cases. Here we compare California, Florida, and New York during the first months represented in the data. Weekly aggregation smooths daily fluctuations and makes the state trajectories easier to compare.

```{r}
#| label: prepare-selected-states

selected_states <- google_mobility_US |>
  filter(
    sub_region_1 %in% c("California", "Florida", "New York"),
    date <= as.Date("2020-06-30")
  ) |>
  mutate(week = floor_date(date, "week")) |>
  group_by(sub_region_1, week) |>
  summarise(
    retail = mean(
      retail_and_recreation_percent_change_from_baseline,
      na.rm = TRUE
    ),
    parks = mean(parks_percent_change_from_baseline, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
#| label: fig-selected-state-retail
#| fig-cap: "Weekly retail and recreation mobility in three selected states during early 2020."

ggplot(selected_states, aes(week, retail, color = sub_region_1)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey65") +
  geom_line(linewidth = 0.9) +
  facet_wrap(~sub_region_1, ncol = 1) +
  scale_color_brewer(palette = "Set1", guide = "none") +
  labs(x = NULL, y = "Retail & recreation change (%)")
```

The small multiples use the same vertical scale, so differences in depth and recovery are visually meaningful. Faceting also prevents one state's line from obscuring another.

## Combine multiple plots

The `cowplot` package can assemble separately constructed ggplot objects into one figure. This is useful when outcomes have different ranges or labels but should be read together.

```{r}
#| label: fig-selected-state-comparison
#| fig-cap: "Retail and parks mobility for California, Florida, and New York during early 2020."
#| fig-height: 8

retail_plot <- ggplot(selected_states, aes(week, retail, color = sub_region_1)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey65") +
  geom_line(linewidth = 0.85) +
  scale_color_brewer(palette = "Set1") +
  labs(x = NULL, y = "Retail & recreation (%)", color = "State") +
  theme(legend.position = "bottom")

parks_plot <- ggplot(selected_states, aes(week, parks, color = sub_region_1)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey65") +
  geom_line(linewidth = 0.85) +
  scale_color_brewer(palette = "Set1") +
  labs(x = NULL, y = "Parks (%)", color = "State") +
  theme(legend.position = "bottom")

plot_grid(retail_plot, parks_plot, ncol = 1, align = "v")
```

Notice that parks mobility behaves differently from retail activity. Treating every mobility category as interchangeable would erase those differences, so the outcome should always be chosen to match the research question.

## Evaluate the largest changes

Finally, compare each state's average parks mobility in 2020 and 2021. Reshaping the yearly values into columns makes the difference easy to compute. This is a descriptive comparison, not evidence that the passage of one year caused the change.

```{r}
#| label: compute-largest-changes

parks_change <- google_mobility_US |>
  filter(dt_year %in% c(2020, 2021)) |>
  group_by(sub_region_1, dt_year) |>
  summarise(
    avg_change_parks = mean(
      parks_percent_change_from_baseline,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from = dt_year,
    values_from = avg_change_parks,
    names_prefix = "year_"
  ) |>
  mutate(change_2021_vs_2020 = year_2021 - year_2020) |>
  arrange(desc(change_2021_vs_2020))
```

```{r}
#| label: tbl-largest-changes
#| tbl-cap: "States with the largest increases in average parks mobility from 2020 to 2021."

parks_change |>
  slice_head(n = 10) |>
  transmute(
    State = sub_region_1,
    `2020 average` = round(year_2020, 1),
    `2021 average` = round(year_2021, 1),
    Difference = round(change_2021_vs_2020, 1)
  ) |>
  kable()
```

```{r}
#| label: fig-largest-changes
#| fig-cap: "States with the largest increases in parks mobility between 2020 and 2021."

parks_change |>
  slice_head(n = 10) |>
  mutate(sub_region_1 = fct_reorder(sub_region_1, change_2021_vs_2020)) |>
  ggplot(aes(change_2021_vs_2020, sub_region_1)) +
  geom_col(fill = "#5b4ee8", width = 0.72) +
  labs(
    x = "Change in average mobility (percentage points)",
    y = NULL
  )
```

The table supports exact lookup, while the bar chart makes ranking and magnitude easier to perceive. Providing both is often more useful than forcing one representation to do both jobs.

## What this workflow demonstrates

This example combines several reusable data-science techniques:

- reading a documented local data snapshot;
- checking dimensions and date coverage before analysis;
- joining observations to a lookup table with an explicit key;
- reshaping between wide and long formats;
- aggregating daily observations to monthly or weekly summaries;
- reducing overplotting through grouping, filtering, and faceting; and
- presenting compact tables alongside publication-ready visualizations.

Most importantly, the rendered results are generated directly from the displayed R code whenever the site is built. The explanation around each chunk tells readers not only **what** the code does, but also **why** the transformation or visualization is appropriate.
